---
title: "Simulation Challenge"
author: "Hailey Pease"
format: 
  html: default
  pdf: 
    documentclass: article
    geometry: margin=1in
    fig-width: 10
    fig-height: 6
    fig-dpi: 300
execute:
  echo: false
  eval: true
  warning: false
  message: false
---

# Investment Game Simulation Analysis

This report analyzes two different investment strategies using Monte Carlo simulation to understand their risk and return characteristics. We'll examine how small differences in strategy rules can lead to dramatically different outcomes over time.

## Question 1: Expected Value Analysis

Let's start by calculating the expected value of your account balance after just one coin flip for the original game. This tells us what we should expect to happen on average.

**Original Game Rules:**
- Start with $1,000
- Heads: increase balance by 50% (multiply by 1.5)
- Tails: reduce balance by 40% (multiply by 0.6)
- Probability of heads = 0.5, tails = 0.5

**What we're calculating:** Expected value is the average outcome if we repeated this game many times. It's calculated by multiplying each possible outcome by its probability and adding them up.

```{r}
# Expected value calculation after 1 coin flip
initial_balance <- 1000

# Possible outcomes after 1 flip
heads_outcome <- initial_balance * 1.5  # 50% gain
tails_outcome <- initial_balance * 0.6  # 40% loss

# Probabilities
prob_heads <- 0.5
prob_tails <- 0.5

# Expected value calculation
expected_value <- (prob_heads * heads_outcome) + (prob_tails * tails_outcome)

cat("Initial balance: $", format(initial_balance, big.mark = ","), "\n", sep = "")
cat("Balance after heads (50% gain): $", format(heads_outcome, big.mark = ","), "\n", sep = "")
cat("Balance after tails (40% loss): $", format(tails_outcome, big.mark = ","), "\n", sep = "")
cat("Expected value after 1 flip: $", format(expected_value, digits = 2, big.mark = ","), "\n", sep = "")
```

**Key Insight:** The expected value of $1,050 means that on average, we expect to gain $50 per flip. This is positive, suggesting the game is favorable in the long run.

## Question 2: Expectation vs. Reality Analysis

Now let's understand what this expected value actually means for our investment strategy.

```{r}
# Analysis of expectation vs. reality
change_from_initial <- expected_value - initial_balance
percentage_change <- (change_from_initial / initial_balance) * 100

cat("\nExpected value: $", format(expected_value, digits = 2, big.mark = ","), "\n", sep = "")
cat("Change from initial: $", format(change_from_initial, digits = 2, big.mark = ","), "\n", sep = "")
cat("Percentage change: ", format(percentage_change, digits = 2), "%\n", sep = "")

if (expected_value > initial_balance) {
  cat("The expected value is POSITIVE - we expect to gain money on average\n")
} else {
  cat("The expected value is NEGATIVE - we expect to lose money on average\n")
}

cat("\nBased on this result, we expect our account to be worth $", format(expected_value, digits = 2, big.mark = ","), "\n", sep = "")
if (change_from_initial > 0) {
  cat("This is $", format(abs(change_from_initial), digits = 2, big.mark = ","), " MORE than our initial $1,000\n", sep = "")
} else {
  cat("This is $", format(abs(change_from_initial), digits = 2, big.mark = ","), " LESS than our initial $1,000\n", sep = "")
}
```

**Key Insight:** Even though we expect to gain money on average (+5%), this doesn't guarantee we'll make money in any single game. The expected value tells us about long-term trends, but individual outcomes can vary widely.

## Question 3: Single Simulation with Time Series Plot

Now let's see what happens when we actually play this game over 30 years (from age 25 to 55). This simulation shows one possible path your investment could take.

**What we're doing:** We'll flip a coin 30 times, each time either gaining 50% or losing 40% of our current balance. This shows how compounding works - both gains and losses build on each other over time.

```{r}
# Load required libraries
library(tidyverse)
library(ggplot2)

# Set seed for reproducibility
set.seed(123)

# Parameters
initial_balance <- 1000
years <- 30  # From age 25 to 55
ages <- 25:55  # Ages 25 through 55

# Simulate one path
simulate_investment_path <- function(initial, years) {
  balance <- initial
  path <- numeric(years + 1)
  path[1] <- initial
  
  for (year in 1:years) {
    coin_flip <- rbinom(1, 1, 0.5)
    if (coin_flip == 1) {
      balance <- balance * 1.5  # 50% gain
    } else {
      balance <- balance * 0.6  # 40% loss
    }
    path[year + 1] <- balance
  }
  
  return(path)
}

# Run single simulation
balance_path <- simulate_investment_path(initial_balance, years)

# Create DataFrame for plotting
simulation_data <- tibble(
  age = ages,
  balance = balance_path
)

# Expected value after 30 years
expected_after_30 <- initial_balance * ((1.5 * 0.5) + (0.6 * 0.5))^30

# Create time series plot using ggplot2
ggplot(simulation_data, aes(x = age, y = balance)) +
  geom_line(color = "darkblue", linewidth = 1.2) +
  geom_point(color = "darkblue", size = 2) +
  geom_hline(yintercept = initial_balance, color = "red", linetype = "dashed", 
             alpha = 0.7, linewidth = 1) +
  geom_hline(yintercept = expected_after_30, color = "green", linetype = "dotted", 
             alpha = 0.7, linewidth = 1) +
  labs(
    title = "Account Balance Over Time - Single Simulation",
    subtitle = "Investment Game from Age 25 to 55",
    x = "Age",
    y = "Account Balance ($)",
    caption = paste("Red line: Initial Balance ($", format(initial_balance, big.mark = ","), 
                   "), Green line: Expected Value After 30 Years ($", 
                   format(expected_after_30, digits = 1, big.mark = ","), ")", sep = "")
  ) +
  scale_y_continuous(labels = scales::dollar_format()) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 14, color = "gray50"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  ) +
  ylim(0, max(balance_path) * 1.1)

# Display the simulation results
cat("Single Simulation Results:\n")
cat(paste(rep("=", 50), collapse = ""), "\n")
cat("Initial balance (age 25): $", format(initial_balance, big.mark = ","), "\n", sep = "")
cat("Final balance (age 55): $", format(balance_path[length(balance_path)], digits = 2, big.mark = ","), "\n", sep = "")
cat("Total change: $", format(balance_path[length(balance_path)] - initial_balance, digits = 2, big.mark = ","), "\n", sep = "")
cat("Total return: ", format(((balance_path[length(balance_path)] / initial_balance) - 1) * 100, digits = 2), "%\n", sep = "")

# Count heads vs tails in this simulation
coin_flips <- character(years)
for (i in 1:years) {
  if (balance_path[i+1] / balance_path[i] == 1.5) {
    coin_flips[i] <- "Heads"
  } else {
    coin_flips[i] <- "Tails"
  }
}

heads_count <- sum(coin_flips == "Heads")
tails_count <- sum(coin_flips == "Tails")
cat("\nCoin flip results: ", heads_count, " heads, ", tails_count, " tails\n", sep = "")
cat("Heads percentage: ", format((heads_count/length(coin_flips))*100, digits = 1), "%\n", sep = "")
```

**What this plot shows:** The blue line shows how your account balance changes over time. The red dashed line shows your starting point ($1,000), and the green dotted line shows the expected value after 30 years. Notice how the actual path can be very different from the expected value due to random chance.

## Analysis of Single Simulation Results

```{r}
# Analysis and commentary on the single simulation
final_balance <- balance_path[length(balance_path)]

cat("\n", paste(rep("=", 60), collapse = ""), "\n")
cat("ANALYSIS OF SINGLE SIMULATION\n")
cat(paste(rep("=", 60), collapse = ""), "\n")

cat("\n🎯 Key Results:\n")
cat("   • Started with: $", format(initial_balance, big.mark = ","), "\n", sep = "")
cat("   • Ended with: $", format(final_balance, digits = 2, big.mark = ","), "\n", sep = "")
cat("   • Net change: $", format(final_balance - initial_balance, digits = 2, big.mark = ","), "\n", sep = "")

if (final_balance > initial_balance) {
  cat("   • This simulation was PROFITABLE! 🎉\n")
  cat("   • You would be happy with this outcome\n")
} else {
  cat("   • This simulation resulted in a LOSS 😞\n")
  cat("   • You would NOT be happy with this outcome\n")
}

cat("\n📊 Performance Analysis:\n")
total_return <- ((final_balance / initial_balance) - 1) * 100
annualized_return <- (((final_balance / initial_balance) ^ (1/30)) - 1) * 100
cat("   • Total return: ", format(total_return, digits = 2), "%\n", sep = "")
cat("   • Annualized return: ", format(annualized_return, digits = 2), "%\n", sep = "")

cat("\n🎲 Luck Factor:\n")
expected_heads <- years * 0.5
cat("   • Got ", heads_count, " heads out of ", years, " flips\n", sep = "")
cat("   • Expected heads: ", format(expected_heads, digits = 1), "\n", sep = "")
if (heads_count > expected_heads) {
  cat("   • You were LUCKY - got more heads than expected! 🍀\n")
} else if (heads_count < expected_heads) {
  cat("   • You were UNLUCKY - got fewer heads than expected! 😔\n")
} else {
  cat("   • You got exactly the expected number of heads\n")
}

cat("\n💡 Key Insight:\n")
cat("   • Even though the expected value per flip is positive ($", format(expected_value, digits = 2, big.mark = ","), ")\n", sep = "")
cat("   • Individual outcomes can vary dramatically due to compounding\n")
cat("   • This single simulation shows the high variability possible\n")
cat("   • Multiple simulations would give a better picture of the true risk\n")
```

**Key Insight:** One simulation doesn't tell us the whole story. To understand the true risk and potential outcomes, we need to run many simulations and look at the distribution of results.

## Question 4: Multiple Simulations (100 simulations)

Now let's run 100 different simulations to see the range of possible outcomes. This will show us the distribution of final balances and help us understand the true risk of this investment strategy.

**What we're doing:** We'll run the same 30-year simulation 100 times with different random coin flips. This gives us a probability distribution showing all possible outcomes and their likelihoods.

```{r}
# Set seed for reproducibility
set.seed(456)

# Parameters for multiple simulations
n_sims <- 100

# Function to simulate final balance only
simulate_final_balance <- function(initial, years) {
  balance <- initial
  for (year in 1:years) {
    coin_flip <- rbinom(1, 1, 0.5)
    if (coin_flip == 1) {
      balance <- balance * 1.5  # 50% gain
    } else {
      balance <- balance * 0.6  # 40% loss
    }
  }
  return(balance)
}

# Run 100 simulations
final_balances <- replicate(n_sims, simulate_final_balance(initial_balance, years))

# Create data frame for analysis
distribution_data <- tibble(
  sim_num = 1:n_sims,
  final_balance = final_balances
)

# Create histogram of final balances
ggplot(distribution_data, aes(x = final_balance)) +
  geom_histogram(bins = 20, fill = "plum", alpha = 0.8, color = "black") +
  geom_vline(xintercept = initial_balance, color = "red", linetype = "dashed", linewidth = 1.5) +
  geom_vline(xintercept = expected_after_30, color = "green", linetype = "dotted", linewidth = 1.5) +
  labs(
    title = "Distribution of Final Account Balances",
    subtitle = paste("100 Simulations,", years, "Years Each"),
    x = "Final Balance ($)",
    y = "Frequency",
    caption = paste("Red line: Initial Balance ($", format(initial_balance, big.mark = ","), 
                   "), Green line: Expected Value ($", 
                   format(expected_after_30, digits = 1, big.mark = ","), ")", sep = "")
  ) +
  scale_x_continuous(labels = scales::dollar_format()) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 14, color = "gray50"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )

# Summary statistics
summary_stats <- distribution_data %>%
  summarise(
    mean_balance = mean(final_balance),
    median_balance = median(final_balance),
    min_balance = min(final_balance),
    max_balance = max(final_balance),
    prob_above_initial = mean(final_balance > initial_balance),
    prob_above_10000 = mean(final_balance > 10000)
  )

cat("Summary Statistics for 100 Simulations:\n")
cat(paste(rep("=", 50), collapse = ""), "\n")
cat("Mean final balance: $", format(summary_stats$mean_balance, digits = 2, big.mark = ","), "\n", sep = "")
cat("Median final balance: $", format(summary_stats$median_balance, digits = 2, big.mark = ","), "\n", sep = "")
cat("Minimum final balance: $", format(summary_stats$min_balance, digits = 2, big.mark = ","), "\n", sep = "")
cat("Maximum final balance: $", format(summary_stats$max_balance, digits = 2, big.mark = ","), "\n", sep = "")
cat("Probability above initial ($1,000): ", format(summary_stats$prob_above_initial, digits = 3), " (", format(summary_stats$prob_above_initial * 100, digits = 1), "%)\n", sep = "")
cat("Probability above $10,000: ", format(summary_stats$prob_above_10000, digits = 3), " (", format(summary_stats$prob_above_10000 * 100, digits = 1), "%)\n", sep = "")

# Analysis
cat("\n📊 Analysis of 100 Simulations:\n")
if (summary_stats$prob_above_initial > 0.5) {
  cat("   • You would be happy with ", format(summary_stats$prob_above_initial * 100, digits = 1), "% of outcomes\n", sep = "")
} else {
  cat("   • You would NOT be happy with ", format((1 - summary_stats$prob_above_initial) * 100, digits = 1), "% of outcomes\n", sep = "")
}

if (summary_stats$median_balance > initial_balance) {
  cat("   • The median outcome is profitable\n")
} else {
  cat("   • The median outcome results in a loss\n")
}

cat("   • There's huge variability: from $", format(summary_stats$min_balance, digits = 1, big.mark = ","), 
    " to $", format(summary_stats$max_balance, digits = 1, big.mark = ","), "\n", sep = "")
```

## Question 5: Probability Analysis

```{r}
# Probability analysis based on 100 simulations
prob_above_1000 <- summary_stats$prob_above_initial

cat("\n", paste(rep("=", 60), collapse = ""), "\n")
cat("PROBABILITY ANALYSIS - ORIGINAL STRATEGY\n")
cat(paste(rep("=", 60), collapse = ""), "\n")

cat("\n🎯 Key Probability Results:\n")
cat("   • Probability that balance > $1,000 at age 55: ", format(prob_above_1000, digits = 3), 
    " (", format(prob_above_1000 * 100, digits = 1), "%)\n", sep = "")
cat("   • Probability that balance ≤ $1,000 at age 55: ", format(1 - prob_above_1000, digits = 3), 
    " (", format((1 - prob_above_1000) * 100, digits = 1), "%)\n", sep = "")

cat("\n💡 Interpretation:\n")
if (prob_above_1000 > 0.5) {
  cat("   • You have a BETTER than 50% chance of ending up with more than you started!\n")
  cat("   • This suggests the strategy is favorable in the long run\n")
} else {
  cat("   • You have a LESS than 50% chance of ending up with more than you started!\n")
  cat("   • This suggests the strategy is unfavorable in the long run\n")
}

# Count how many simulations were above/below initial
above_1000_count <- sum(final_balances > initial_balance)
below_1000_count <- sum(final_balances <= initial_balance)

cat("\n📈 Detailed Breakdown:\n")
cat("   • ", above_1000_count, " out of ", n_sims, " simulations ended above $1,000\n", sep = "")
cat("   • ", below_1000_count, " out of ", n_sims, " simulations ended at or below $1,000\n", sep = "")
```

**Key Insight:** The histogram shows the distribution of final balances across 100 simulations. Most outcomes cluster around certain values, but there's a wide range of possible results. This is the reality of investment risk - even with a positive expected value, individual outcomes can vary dramatically.

## Question 6: Strategy Comparison - Modified Game

Now let's compare the original strategy with a modified version to see how different rules affect outcomes.

**Modified Game Rules:**
- Start with $1,000
- Must bet exactly 50% of current balance each round
- Heads: increase your bet by 50% (add 50% of bet to balance)
- Tails: reduce your bet by 40% (subtract 40% of bet from balance)
- Play for 30 years (age 25 to 55)

**Key Difference:** In the modified game, you can only bet 50% of your balance each round, rather than betting your entire balance. This should theoretically reduce risk, but let's see what the simulations show.

```{r}
# Set seed for reproducibility
set.seed(789)

# Function to simulate modified game
simulate_modified_game <- function(initial, years) {
  balance <- initial
  for (year in 1:years) {
    bet_amount <- balance * 0.5  # Bet exactly 50% of balance
    coin_flip <- rbinom(1, 1, 0.5)
    if (coin_flip == 1) {
      # Heads: increase bet by 50%
      balance <- balance + (bet_amount * 0.5)
    } else {
      # Tails: reduce bet by 40%
      balance <- balance - (bet_amount * 0.4)
    }
  }
  return(balance)
}

# Run 100 simulations for modified game
modified_final_balances <- replicate(n_sims, simulate_modified_game(initial_balance, years))

# Create data frame for modified game
modified_distribution_data <- tibble(
  sim_num = 1:n_sims,
  final_balance = modified_final_balances
)

# Summary statistics for modified game
modified_summary_stats <- modified_distribution_data %>%
  summarise(
    mean_balance = mean(final_balance),
    median_balance = median(final_balance),
    min_balance = min(final_balance),
    max_balance = max(final_balance),
    prob_above_initial = mean(final_balance > initial_balance),
    prob_above_10000 = mean(final_balance > 10000)
  )

cat("\n", paste(rep("=", 60), collapse = ""), "\n")
cat("MODIFIED GAME STRATEGY RESULTS\n")
cat(paste(rep("=", 60), collapse = ""), "\n")

cat("Summary Statistics for Modified Game (100 Simulations):\n")
cat("Mean final balance: $", format(modified_summary_stats$mean_balance, digits = 2, big.mark = ","), "\n", sep = "")
cat("Median final balance: $", format(modified_summary_stats$median_balance, digits = 2, big.mark = ","), "\n", sep = "")
cat("Minimum final balance: $", format(modified_summary_stats$min_balance, digits = 2, big.mark = ","), "\n", sep = "")
cat("Maximum final balance: $", format(modified_summary_stats$max_balance, digits = 2, big.mark = ","), "\n", sep = "")
cat("Probability above initial ($1,000): ", format(modified_summary_stats$prob_above_initial, digits = 3), 
    " (", format(modified_summary_stats$prob_above_initial * 100, digits = 1), "%)\n", sep = "")
cat("Probability above $10,000: ", format(modified_summary_stats$prob_above_10000, digits = 3), 
    " (", format(modified_summary_stats$prob_above_10000 * 100, digits = 1), "%)\n", sep = "")

# Strategy Comparison
cat("\n", paste(rep("=", 60), collapse = ""), "\n")
cat("STRATEGY COMPARISON ANALYSIS\n")
cat(paste(rep("=", 60), collapse = ""), "\n")

cat("\n🎯 Key Comparison Results:\n")
cat("   • Original Strategy - Probability above $10,000: ", format(summary_stats$prob_above_10000, digits = 3), 
    " (", format(summary_stats$prob_above_10000 * 100, digits = 1), "%)\n", sep = "")
cat("   • Modified Strategy - Probability above $10,000: ", format(modified_summary_stats$prob_above_10000, digits = 3), 
    " (", format(modified_summary_stats$prob_above_10000 * 100, digits = 1), "%)\n", sep = "")

if (modified_summary_stats$prob_above_10000 > summary_stats$prob_above_10000) {
  cat("   • The MODIFIED strategy has a HIGHER probability of reaching $10,000! 🎉\n")
} else if (modified_summary_stats$prob_above_10000 < summary_stats$prob_above_10000) {
  cat("   • The ORIGINAL strategy has a HIGHER probability of reaching $10,000! 🎯\n")
} else {
  cat("   • Both strategies have the SAME probability of reaching $10,000! 🤔\n")
}

difference <- modified_summary_stats$prob_above_10000 - summary_stats$prob_above_10000
cat("   • Difference: ", format(difference, digits = 3), " (", format(difference * 100, digits = 1), "%)\n", sep = "")

cat("\n💡 Strategic Insights:\n")
if (modified_summary_stats$mean_balance > summary_stats$mean_balance) {
  cat("   • Modified strategy has higher average returns\n")
} else {
  cat("   • Original strategy has higher average returns\n")
}

if (modified_summary_stats$median_balance > summary_stats$median_balance) {
  cat("   • Modified strategy has higher median returns\n")
} else {
  cat("   • Original strategy has higher median returns\n")
}

cat("   • Modified strategy range: $", format(modified_summary_stats$min_balance, digits = 1, big.mark = ","), 
    " to $", format(modified_summary_stats$max_balance, digits = 1, big.mark = ","), "\n", sep = "")
cat("   • Original strategy range: $", format(summary_stats$min_balance, digits = 1, big.mark = ","), 
    " to $", format(summary_stats$max_balance, digits = 1, big.mark = ","), "\n", sep = "")

# Create comparison plot
comparison_data <- tibble(
  strategy = rep(c("Original", "Modified"), each = n_sims),
  final_balance = c(final_balances, modified_final_balances)
)

ggplot(comparison_data, aes(x = final_balance, fill = strategy)) +
  geom_histogram(alpha = 0.7, position = "identity", bins = 20) +
  geom_vline(xintercept = 10000, color = "red", linetype = "dashed", linewidth = 1.5) +
  labs(
    title = "Strategy Comparison: Final Balance Distributions",
    subtitle = "100 Simulations Each - Original vs Modified Game",
    x = "Final Balance ($)",
    y = "Frequency",
    fill = "Strategy",
    caption = "Red line: $10,000 threshold"
  ) +
  scale_x_continuous(labels = scales::dollar_format()) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    plot.subtitle = element_text(size = 14, color = "gray50"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.position = "bottom"
  ) +
  facet_wrap(~strategy, ncol = 1, scales = "free_y")
```

## Final Analysis and Recommendations

```{r}
cat("\n", paste(rep("=", 70), collapse = ""), "\n")
cat("FINAL ANALYSIS AND RECOMMENDATIONS\n")
cat(paste(rep("=", 70), collapse = ""), "\n")

cat("\n🎲 Key Findings:\n")
cat("1. Expected value per flip is positive ($", format(expected_value, digits = 2, big.mark = ","), ")\n", sep = "")
cat("2. Original strategy: ", format(summary_stats$prob_above_10000 * 100, digits = 1), "% chance of reaching $10,000\n", sep = "")
cat("3. Modified strategy: ", format(modified_summary_stats$prob_above_10000 * 100, digits = 1), "% chance of reaching $10,000\n", sep = "")

if (modified_summary_stats$prob_above_10000 > summary_stats$prob_above_10000) {
  cat("4. The MODIFIED strategy is better for reaching $10,000\n")
} else {
  cat("4. The ORIGINAL strategy is better for reaching $10,000\n")
}

cat("\n💡 Investment Implications:\n")
cat("   • Both strategies show the power of compounding over time\n")
cat("   • High variability means outcomes can be dramatically different\n")
cat("   • The 'safer' modified strategy (betting only 50%) may actually be riskier\n")
cat("   • Individual luck plays a huge role in short-term outcomes\n")
cat("   • Multiple simulations reveal the true risk profile\n")

cat("\n🎯 Recommendation for a 25-year-old:\n")
if (modified_summary_stats$prob_above_10000 > summary_stats$prob_above_10000) {
  cat("   Consider the MODIFIED strategy for higher probability of reaching $10,000\n")
} else {
  cat("   Consider the ORIGINAL strategy for higher probability of reaching $10,000\n")
}

cat("   However, remember that past performance doesn't guarantee future results!\n")
cat("   Diversification and professional financial advice are always recommended.\n")
```

